
\def\doctitle{{\tt ensmallen}: accelerating mathematical optimization via template metaprogramming}
\def\docauthor{AuthorA, Author B, Author C, ...}

%% document title and authors are defined above,
%% in order to embed this info into PDF metadata via the hyperref package

\documentclass[10pt]{article}
\usepackage[latin1]{inputenc}
\usepackage[english]{babel}
\usepackage[none]{hyphenat}
\sloppy  %% this actually gives better linebreaks, minimising overhanging text
\usepackage[T1]{fontenc}
%\usepackage{url}
\usepackage[includefoot,includehead,left=2cm,right=2cm,top=1cm,bottom=2cm]{geometry}
\usepackage[usenames,dvipsnames]{color}  % see http://en.wikibooks.org/wiki/LaTeX/Colors
\usepackage[pdfborder={0 0 0},colorlinks=true,urlcolor=blue,linkcolor=blue,citecolor=blue,bookmarks=true,pdftitle={\doctitle},pdfauthor={\docauthor}]{hyperref}
\usepackage[numbers]{natbib}
\usepackage{graphicx}
%\usepackage[draft]{minted}
\usepackage{minted}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{booktabs}      % for \toprule in tables
\usepackage{multirow}      % for table multirow
\usepackage{tcolorbox}   % colorbox in minted
\usepackage{times}
%\usepackage{palatino}     % changes the default sans serif font
\usepackage{inconsolata}   % changes the default typewriter font
\usepackage{adjustbox}
\usepackage{threeparttable}
\usepackage{parcolumns}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{tabularx}
\usepackage{array}
\usepackage{wasysym}

% disable ligatures such as "fi" being joined into one symbol
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

\renewcommand{\baselinestretch}{1.1}\small\normalsize

\newcolumntype{R}[2]{%
  >{\adjustbox{angle=#1,lap=\width-(#2)}\bgroup}%
  l%
  <{\egroup}%
}
\newcommand*\rot{\multicolumn{1}{R{30}{2.0em}}}

\def\TODO#1{{\color{red}{\bf [TODO:} {{#1}}{\bf ]}}}

\begin{document}

%% see the \doctitle  definition at the top
%% see the \docauthor definition at the top

\begin{center}
{\Large\doctitle}
\end{center}
\vspace{1ex}
\begin{center}
{\docauthor}
\end{center}

\vspace{5ex} % space for authors/affiliations

\section*{Abstract}
This paper provides an introduction to the {\tt ensmallen} numerical optimization library,
as well as a deep dive into the technical details of how it works.
The library provides a fast and flexible C++ framework
for mathematical optimization of arbitrary user-supplied functions.
A large set of pre-built optimizers is provided,
including many variants of SGD and Quasi-Newton optimizers.
Several types of objective functions are supported, including differentiable,
separable, constrained, and categorical objective functions.
Implementation of a new optimizer requires only one method,
while a new objective function requires typically only one or two C++ methods.
Through C++ template metaprogramming, {\tt ensmallen} provides support for arbitrary
user-supplied callbacks and automatic inference of unsupplied methods without
any runtime overhead.
Empirical comparisons show that {\tt ensmallen}
outperforms other optimization frameworks (such as Julia and SciPy), sometimes
by large margins.  The library is available at \url{https://ensmallen.org}
and is distributed under the permissive BSD license.


% and is ready for use in productions environments.  
%% Commented out, as it's a subjective statement;
%% whether it's "ready" for a specific "production environment"
%% is up to the user.
%% Stating the BSD license is already a big hint that
%% people can do whatever they want with it,
%% and that there are no guarantees of fitness
%% for a particular purpose.
%% https://tldrlegal.com/license/bsd-3-clause-license-(revised)#fulltext


\vspace{4ex}
\hrule


\input{sections/introduction.tex}
\input{sections/api.tex}
\input{sections/templated_optimize.tex}
\input{sections/functions.tex}
\input{sections/callbacks.tex}
\input{sections/experiments.tex}
\input{sections/conclusion.tex}

\bibliographystyle{ieee}
\bibliography{refs}

\clearpage
% Could use \appendix here if we wanted, but I figured a \clearpage was enough.
\input{sections/templated_optimize_details.tex}
\input{sections/automatic_details.tex}
\input{sections/callbacks_internals.tex}

\end{document}
