\section*{Abstract}

%\begin{small}
This report provides an introduction to the {\tt ensmallen} numerical optimization library,
as well as a deep dive into the technical details of how it works.
The library provides a fast and flexible C++ framework
for mathematical optimization of arbitrary user-supplied functions.
A large set of pre-built optimizers is provided,
including many variants of Stochastic Gradient Descent and Quasi-Newton optimizers.
Several types of objective functions are supported, including differentiable,
separable, constrained, and categorical objective functions.
Implementation of a new optimizer requires only one method,
while a new objective function requires typically only one or two C++ methods.
Through internal use of C++ template metaprogramming, {\tt ensmallen} provides support for arbitrary
user-supplied callbacks and automatic inference of unsupplied methods without
any runtime overhead.
Empirical comparisons show that {\tt ensmallen}
outperforms other optimization frameworks (such as Julia and SciPy), sometimes
by large margins.  The library is available at \url{https://ensmallen.org}
and is distributed under the permissive BSD license.
%\end{small}

% and is ready for use in productions environments.  
%% Commented out, as it's a subjective statement;
%% whether it's "ready" for a specific "production environment"
%% is up to the user.
%% Stating the BSD license is already a big hint that
%% people can do whatever they want with it,
%% and that there are no guarantees of fitness
%% for a particular purpose.
%% https://tldrlegal.com/license/bsd-3-clause-license-(revised)#fulltext


